---
title: "STA 652 Assignment One - Richel Attafuah"
author: "Richel Attafuah"
date: "2026-03-16"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Loading all libraries
library(readr)
library(dplyr)
library(lubridate)
library(purrr)
library(tidyr)
library(ggplot2)
library(forecast)
```

# Introduction

Time series forecasting plays a critical role in understanding and predicting patterns in real-world systems. Many real-world time series exhibit complex seasonal structures, where patterns repeat at multiple time scales rather than a single periodic cycle. Traditional models such as Seasonal Autoregressive Integrated Moving Average (SARIMA) are designed to capture only one seasonal pattern, which limits their effectiveness when multiple seasonalities are present.

This study investigates the forecasting of urban mobility demand using real-world bike-sharing data from New York City's Citi Bike system. Bike usage patterns are influenced by daily commuting behavior as well as weekly variations between weekdays and weekends, resulting in multiple seasonal cycles.

The objectives of this analysis are threefold. First, to explore and identify the seasonal patterns present in the data. Second, to evaluate the performance of SARIMA models using both manual identification and automatic selection. Third, to compare these models with alternative approaches specifically designed to handle multiple seasonal patterns, including Fourier-based methods and TBATS models.

By systematically comparing these approaches, this study aims to demonstrate the limitations of single-seasonal models and highlight the advantages of models capable of capturing multiple seasonal structures.

## Data Preparation

The Citi Bike dataset is provided as multiple files per month, where each file represents a portion of the full monthly data. Therefore, all files within each month were combined to reconstruct complete monthly datasets before aggregation.

Trip-level data was then converted into an hourly time series by counting the number of trips starting in each hour.


```{r echo=TRUE}
files <- list.files(pattern = "\\.csv$", full.names = TRUE)
head(files)
```

```{r echo=TRUE}
july_files <- files[grepl("202507", files)]
july_files
```

```{r echo=TRUE}
july_data <- july_files %>%
  map_dfr(~ read_csv(.x, col_types = cols(.default = "c")))

head(july_data)
```

```{r echo=TRUE}
july_hourly <- july_data %>%
  mutate(
    started_at = ymd_hms(started_at),
    hour = floor_date(started_at, "hour")
  ) %>%
  count(hour, name = "demand") %>%
  complete(
    hour = seq(min(hour), max(hour), by = "hour"),
    fill = list(demand = 0)
  ) %>%
  arrange(hour)

head(july_data)
```



```{r echo=TRUE}
aug_files <- files[grepl("202508", files)]

aug_data <- aug_files %>%
  map_dfr(~ read_csv(.x, col_types = cols(.default = "c")))

aug_hourly <- aug_data %>%
  mutate(
    started_at = ymd_hms(started_at),
    hour = floor_date(started_at, "hour")
  ) %>%
  count(hour, name = "demand") %>%
  complete(
    hour = seq(min(hour), max(hour), by = "hour"),
    fill = list(demand = 0)
  ) %>%
  arrange(hour)

```


```{r echo=TRUE}
sep_files <- files[grepl("202509", files)]

sep_data <- sep_files %>%
  map_dfr(~ read_csv(.x, col_types = cols(.default = "c")))

sep_hourly <- sep_data %>%
  mutate(
    started_at = ymd_hms(started_at),
    hour = floor_date(started_at, "hour")
  ) %>%
  count(hour, name = "demand") %>%
  complete(
    hour = seq(min(hour), max(hour), by = "hour"),
    fill = list(demand = 0)
  ) %>%
  arrange(hour)

```

```{r echo=TRUE}

bike_hourly <- bind_rows(july_hourly, aug_hourly, sep_hourly) %>%
  arrange(hour)

```


```{r echo=TRUE}
nrow(bike_hourly)
min(bike_hourly$hour)
max(bike_hourly$hour)
```
```{r echo=TRUE}
bike_hourly <- bike_hourly %>%
  filter(
    hour >= ymd_h("2025-07-01 00"),
    hour <= ymd_h("2025-09-30 23")
  )
```

```{r echo=TRUE}
nrow(bike_hourly)
min(bike_hourly$hour)
max(bike_hourly$hour)
```

# Exploratory Data Analysis (EDA)

Exploratory Data Analysis (EDA) was conducted to understand the underlying structure of the hourly bike demand time series before applying forecasting models. This step is essential in time series analysis, as it allows for the identification of key patterns such as trends, seasonal cycles, and variability in the data.

In particular, the analysis focuses on detecting the presence of multiple seasonalities, which are expected in this dataset due to human mobility behavior. Hourly bike demand is likely influenced by daily commuting patterns as well as differences between weekdays and weekends. Visual inspection of the time series at different time scales enables the identification of these recurring patterns and helps determine the appropriate modeling approach.

Understanding these patterns is critical, as traditional models such as SARIMA are limited in their ability to capture more than one seasonal component. Therefore, the insights gained from this exploratory analysis will guide the selection and justification of forecasting models used in subsequent sections.

```{r echo=FALSE}
ggplot(bike_hourly, aes(x = hour, y = demand)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Hourly Citi Bike Demand (July–September 2025)",
    x = "Time",
    y = "Number of Trips"
  ) +
  theme_minimal()

```

The full time series plot reveals clear evidence of strong seasonal structure in the hourly bike demand data. The most prominent pattern is a highly regular sequence of peaks and troughs occurring within each day, indicating a strong daily seasonal cycle. These recurring spikes are consistent with commuting behavior, where demand increases during morning and evening hours.

In addition to the daily pattern, a secondary seasonal structure is visible across the weeks. The magnitude and shape of the daily peaks vary over time, suggesting differences between weekdays and weekends. This indicates the presence of a weekly seasonal cycle, where commuter-driven demand is typically higher and more structured during weekdays, while weekend usage appears more irregular.

There is also a slight upward movement in overall demand from July through September, suggesting a mild increasing trend over the observed period. Furthermore, the variability in demand appears to increase with the level of the series, which indicates that the seasonal effects may be multiplicative rather than purely additive.

Overall, the visual inspection confirms that the dataset exhibits multiple seasonal patterns, specifically daily and weekly cycles. This provides strong motivation for using models that can capture more than one seasonal component.


```{r echo=FALSE}
one_week <- bike_hourly %>%
  filter(
    hour >= ymd_h("2025-07-07 00"),
    hour <= ymd_h("2025-07-13 23")
  )

ggplot(one_week, aes(x = hour, y = demand)) +
  geom_line(color = "darkorange") +
  labs(
    title = "Hourly Bike Demand Over One Week",
    x = "Time",
    y = "Number of Trips"
  ) +
  theme_minimal()
```

The weekly time series plot provides a clearer view of both intra-day and inter-day seasonal patterns in bike demand. Within each day, a consistent pattern of sharp peaks and troughs is observed, confirming the presence of strong daily seasonality driven by commuting behavior.

More importantly, distinct differences are visible across the days of the week. During weekdays, demand exhibits highly structured and pronounced peaks, particularly during typical commuting hours in the morning and evening. This reflects regular work-related travel patterns.

In contrast, weekend demand displays a smoother and less structured profile. The peaks are generally lower and more spread out across the day, indicating a shift from commuter-driven usage to more flexible, leisure-oriented travel behavior.

These differences between weekdays and weekends provide strong evidence of a weekly seasonal cycle. Combined with the daily repetition observed within each day, this confirms that the dataset exhibits multiple seasonal patterns, specifically daily and weekly cycles, which must be accounted for in the modeling process.



```{r echo=FALSE,warning=FALSE}

daily_pattern <- bike_hourly %>%
  mutate(hour_of_day = hour(hour)) %>%
  group_by(hour_of_day) %>%
  summarise(
    avg_demand = mean(demand),
    .groups = "drop"
  )

ggplot(daily_pattern, aes(x = hour_of_day, y = avg_demand)) +
  geom_line(color = "darkgreen", size = 1) +
  labs(
    title = "Average Hourly Bike Demand Across a Day",
    x = "Hour of Day",
    y = "Average Number of Trips"
  ) +
  theme_minimal()
```

The average daily demand profile reveals a clear and consistent intra-day seasonal pattern in bike usage. Demand is lowest during the late night and early morning hours, reflecting minimal travel activity during this period.

Beginning in the early morning, demand rises sharply and reaches a first peak during typical commuting hours, indicating increased usage as individuals travel to work. This is followed by a moderate decline and relative stabilization during midday, suggesting reduced mobility during working hours.

A second and more pronounced peak occurs in the late afternoon and early evening, representing the highest demand of the day. This aligns with return commuting patterns, where individuals travel back from work. After this peak, demand gradually declines into the night.

The presence of two distinct peaks within each day confirms a strong daily seasonal cycle driven by commuting behavior. This pattern is highly consistent across the dataset, reinforcing the importance of incorporating daily seasonality into the forecasting models.


# SARIMA Model Identification Using ACF and PACF

To model the time series using SARIMA, it is necessary to identify appropriate model parameters based on the statistical properties of the data. This process involves examining stationarity and analyzing autocorrelation structures through the Autocorrelation Function (ACF) and Partial Autocorrelation Function (PACF).

Since SARIMA models assume that the input series is stationary, the data must first be transformed to remove trends and seasonal patterns. This is achieved through a combination of regular and seasonal differencing. Once stationarity is achieved, the ACF and PACF plots are used to determine suitable values for the autoregressive (AR) and moving average (MA) components, both for the non-seasonal and seasonal parts of the model.

```{r echo=TRUE}
ts_data <- ts(bike_hourly$demand, frequency = 24)
plot(ts_data, main = "Time Series of Hourly Demand")
```


To remove the trend component, first-order differencing is applied. This step helps stabilize the mean of the series over time.


```{r echo=FALSE}
diff_1 <- diff(ts_data, differences = 1)
plot(diff_1, main = "First Differenced Series")
```

Although first differencing removes the trend, the series still contains strong seasonal patterns. To address this, seasonal differencing is applied with a lag of 24, corresponding to the daily cycle. This removes the repeating daily structure and helps achieve stationarity.


```{r echo=FALSE}
diff_seasonal <- diff(diff_1, lag = 24)
plot(diff_seasonal, main = "Differenced Series (d=1, D=1, s=24)")
```

After applying both regular and seasonal differencing, the resulting series is expected to be approximately stationary. The ACF and PACF plots of this transformed series are then used to identify suitable values for the AR and MA components of the SARIMA model.

The ACF plot helps identify potential moving average terms, while the PACF plot provides insight into autoregressive components. Additionally, significant spikes at seasonal lags, particularly multiples of 24, indicate the presence of seasonal AR or MA terms.

```{r echo=FALSE}
acf(diff_seasonal, main = "ACF of Differenced Series")
pacf(diff_seasonal, main = "PACF of Differenced Series")
```
After applying both first-order differencing and seasonal differencing with a lag of 24, the resulting series appears approximately stationary. The transformed series fluctuates around a constant mean with no visible trend, indicating that the differencing successfully removed both trend and daily seasonal components. This confirms the use of d = 1 and D = 1 for the SARIMA model.

The ACF plot of the differenced series shows a strong spike at lag 1 followed by a gradual decay. This pattern suggests the presence of a moving average component in the model. The absence of a sharp cutoff indicates that a low-order MA term is appropriate, with q = 1 being a reasonable choice.

The PACF plot exhibits a significant spike at lag 1, after which the values taper off without a clear cutoff. This behavior indicates the presence of an autoregressive component, suggesting that a first-order AR term, p = 1, is suitable for the model.

Based on the observed patterns in the ACF and PACF plots, a SARIMA(1,1,1)(1,1,1)[24] model is selected as a candidate model. The non-seasonal parameters (p = 1, d = 1, q = 1) capture short-term dependencies, while the seasonal parameters (P = 1, D = 1, Q = 1) account for the daily seasonal structure with a periodicity of 24 hours.

```{r echo=FALSE}
model_manual <- Arima(
  ts_data,
  order = c(1,1,1),
  seasonal = c(1,1,1)
)

summary(model_manual)

```

The manually specified SARIMA(1,1,1)(1,1,1)[24] model was fitted to the differenced time series. The estimated coefficients for both non-seasonal and seasonal components are statistically significant and indicate that the model captures both short-term dependencies and daily seasonal structure.

The autoregressive and moving average terms reflect strong temporal dependence in the series, while the seasonal components capture the repeating daily cycle. The model achieves a reasonable fit based on RMSE and MAE values.

It is noted that the Mean Absolute Percentage Error (MAPE) is not reliable in this case due to the presence of zero demand values, which result in division by zero. Therefore, evaluation focuses on RMSE and MAE as more appropriate performance metrics.

```{r echo=FALSE}
checkresiduals(model_manual)
```

Residual diagnostics were performed to assess the adequacy of the fitted SARIMA(1,1,1)(1,1,1)[24] model. The residual time series appears to fluctuate around zero; however, periods of increased variability suggest that not all patterns in the data have been captured.

The ACF plot of the residuals reveals several significant spikes, indicating the presence of remaining autocorrelation. This suggests that the model has not fully accounted for all temporal dependencies in the series.

The Ljung-Box test further confirms this observation, with a p-value significantly less than 0.05. This leads to the rejection of the null hypothesis of no autocorrelation, indicating that the residuals are not white noise.

Overall, these results suggest that the SARIMA model does not fully capture the underlying structure of the data. This limitation is likely due to the presence of multiple seasonal patterns, which a single-seasonal SARIMA model is not designed to handle.


```{r echo=FALSE}
model_auto <- auto.arima(ts_data)
summary(model_auto)
checkresiduals(model_auto)
```

An automatic SARIMA model was fitted using the auto.arima() function to provide a benchmark for comparison. The selected model, ARIMA(5,0,4)(2,1,0)[24] with drift, includes higher-order autoregressive and moving average terms, as well as a drift component to capture potential trends in the data.

Compared to the manually specified model, the automatic model is more complex and attempts to capture additional dependencies in the time series. This results in a slight improvement in fit, as indicated by lower RMSE and MAE values.

However, residual diagnostics indicate that the model is still inadequate. The residual time series shows remaining patterns, and the ACF plot reveals significant autocorrelations at several lags. The Ljung-Box test produces a p-value far below 0.05, leading to rejection of the null hypothesis of no autocorrelation.

This indicates that the residuals are not white noise, meaning that the model has not fully captured the underlying structure of the data.

Despite the increased complexity of the automatic SARIMA model, it still fails to adequately model the data. This suggests that the limitation is not due to parameter selection, but rather due to the structural inability of SARIMA models to capture multiple seasonal patterns simultaneously.

In this dataset, both daily and weekly seasonal cycles are present. Since SARIMA can only model a single seasonal period at a time, it is unable to fully represent the multi-seasonal nature of the data. This motivates the use of alternative models specifically designed to handle multiple seasonalities.

# Modeling Multiple Seasonalities

## Fourier Terms with ARIMA Errors

To address the limitations of SARIMA in handling multiple seasonal patterns, a Fourier-based approach combined with ARIMA errors is applied. This method extends traditional ARIMA modeling by incorporating Fourier terms as external regressors to capture complex and multiple seasonal cycles.

Fourier terms are constructed using sine and cosine functions at different frequencies, allowing the model to represent periodic patterns more flexibly than standard seasonal differencing. In this case, two seasonal cycles are present in the data: a daily cycle with a period of 24 hours and a weekly cycle with a period of 168 hours.

By including Fourier terms for both seasonalities, the model can simultaneously capture short-term daily fluctuations and longer-term weekly variations. The ARIMA component is then used to model any remaining temporal dependencies not explained by the seasonal terms.

This approach is particularly effective for multi-seasonal time series, as it avoids the structural limitation of SARIMA models, which can only accommodate a single seasonal period.

```{r echo=FALSE}
ts_data_multi <- msts(
  bike_hourly$demand,
  seasonal.periods = c(24, 168)
)

fourier_terms <- fourier(ts_data_multi, K = c(3, 5))

model_fourier <- auto.arima(
  ts_data_multi,
  xreg = fourier_terms,
  seasonal = FALSE
)

summary(model_fourier)
checkresiduals(model_fourier)
```



